Increased productivity using Atomify - a real-time LAMMPS visualizer

### 2018-06-14 IOPLaTeX.tex
Fixed availability
Improved intro and feature summary

\author[cor1]{Anders Hafreager$^{1}$}
\address{$^1$Department of Physics - University of Oslo\\Sem S{\ae}lands vei 24, NO-0316, Oslo, Norway}
\eads{\mailto{anderhaf@fys.uio.no}, \mailto{andershaf@gmail.com}}

\author{Svenn-Arne Dragly$^{1}$}
\ead{s.a.dragly@fys.uio.no}

\begin{abstract}
The typical workflow when running atomistic simulations includes working with several programs.
A text editor is needed to create and modify the scripts, the terminal to run the simulation, and programs like VMD or Ovito to visualize the system over time.
If physical quantities are computed, the data is often plotted using programming languages like MATLAB or Python,
where additional scripts must be used.
This is a tedious process, especially for teaching purposes and for people who are new in the field.
We here introduce Atomify; a high performance real-time visualizer for atomistic simulations that can simulate and render more than one million atoms with excellent frame rate on modern hardware.
Atomify supports OpenMP acceleration, GPU acceleration, real-time plotting of physical quantities, and an easy-to-use code editor in one single application.
It currently uses LAMMPS as physics engine, but it can be extended to support other codes like Gromacs, NAMD or OpenMM.
Atomify is open-source software (GPL) written in C++ using the Qt framework.
\end{abstract}

%Uncomment for PACS numbers title message
\pacs{00.00, 20.00, 42.10}
% Keywords required only for MST, PB, PMB, PM, JOA, JOB? 
\vspace{2pc}
\noindent{\it Keywords}: Article preparation, IOP journals
% Uncomment for Submitted to journal title message
\submitto{\JPA}
% Comment out if separate title page not required
%\maketitle

@@

Over the past decades, users of LAMMPS\cite{Plimpton1995Fast} and other molecular dynamics simulators such as GROMACS, NAMD and CHARMM\cite{berendsen1995gromacs, Phillips2005Scalable, brooks2009charmm} have
typically written scripts in a text editor and run the scripts by passing them to the simulator
on the command line.
The simulator in turn writes the results to disk and the user analyzes the results
by reading the files manually or passing them to other programs.
With the arrival of software like VMD,\cite{Humphrey1996Vmd}
it became easy to both visualize the trajectories and certain 
physical quantities such as the radial distribution function.
VMD came with support both for reading the results from file or attaching to a
running process of supported simulators.
\todo[inline]{Why is Ovito more modern than VMD?}
Later, modern visualization tools like Ovito\cite{Stukowski2009Visualization} has made both visualization and analysis more
accessible to new users through its powerful scriptable modifier pipeline.
In addition, the results can be loaded into common programming languages like MATLAB and Python
to produce publication-ready figures.

@@

It is especially useful during development of new scripts, but can be used
to run full, publication-ready simulations with minimal simulation overhead.
The current implementation only supports LAMMPS, but is written in a general
way so integrating other codes is a simple task.

@@

\section{\label{sec:features}Summary of features}
Atomify was designed to improve the workflow by reducing the number of
context switches and mouse clicks to obtain immediate results from a simulation.
It has a script editor, visualization window and a summary panel so all relevant parts
of a simulation can be on the screen at the same time.

Any changes to the script like defining regions or groups,
testing temperatures, size of simulation box or inserting atoms can
quickly be verified since visualization is real-time.
Rerunning the current script can be done by pressing \keys{Ctrl+R} (\keys{cmd+R} on macOS).

How the pressure, radial distribution function or some other
complicated variable evolves over time is easily understood
since these can be plotted real-time with a simple click.
Bonds between atoms in a molecule is automatically extracted from
the internals of LAMMPS, or can be defined as a distance threshold
between atom types.

The Atomify GUI is shown in figure \ref{fig:gui} where \todo{short summary of GUI}.

This turns out to improve the workflow substantially.
In addition, Atomify comes with multiple examples,
many of which have been made by members of the LAMMPS community.

\section{\label{sec:casestudy}Case study: flow through a narrow nanotube}
The features of Atomify is best demonstrated through a case study where many of
them are useful to obtain a productive workflow during script development.
We have chosen to create a system with fluid flow inside a tight nanochannel.
This involves creating a somewhat complicated initial geometry with multiple regions
where we apply different rules such as a dissipative thermostat, constant force mimicking
applied pressure gradient, and frozen atoms to avoid net momentum of the solid.
We will create two different atom types so we can have a gas phase and a solid phase at the same temperature.

@@

\section{Availability}
Atomify is availabe on Linux, MacOS and Windows and can be downloaded from
\href{https://ovilab.net/atomify}{http://ovilab.net/atomify}.
A lightweight version of Atomify is also available for mobile devices running
Android or iOS.

>>>>

\author[cor1]{Anders Hafreager$^{1}$}
\eads{\mailto{anderhaf@fys.uio.no}, \mailto{andershaf@gmail.com}}

\author{Svenn-Arne Dragly$^{1}$}
\ead{s.a.dragly@fys.uio.no}
\address{$^1$Department of Physics - University of Oslo\\Sem S{\ae}lands vei 24, NO-0316, Oslo, Norway}

\begin{abstract}
The typical workflow when running atomistic simulations includes working with several programs.
A text editor is needed to create and modify the scripts, the terminal to run the simulation, and programs like VMD or Ovito to visualize the system over time.
If physical quantities are computed, the data is often plotted using programming languages like MATLAB or Python,
where additional scripts must be used.
This is a tedious process, especially for teaching purposes and for people who are new in the field.
We here introduce Atomify; a high performance real-time visualizer for atomistic simulations that can simulate and render more than one million atoms with excellent frame rate on modern hardware.
Atomify supports OpenMP acceleration, GPU acceleration, real-time plotting of physical quantities, and an easy-to-use code editor in one single application.
It currently uses LAMMPS as physics engine, but it can be extended to support other codes like GROMACS, NAMD or CHARMM.
Atomify is open-source software (GPL) written in C++ using the Qt framework and can be downloaded from \url{https://github.com/ovilab/atomify}.
\end{abstract}

%Uncomment for PACS numbers title message
\pacs{00.00, 20.00, 42.10}
% Keywords required only for MST, PB, PMB, PM, JOA, JOB? 
\vspace{2pc}
% Comment out if separate title page not required
%\maketitle

@@

Over the past decades, users of LAMMPS\cite{Plimpton1995Fast} and other molecular dynamics simulators such
as GROMACS\cite{berendsen1995gromacs}, NAMD\cite{nelson1996namd} and CHARMM\cite{brooks2009charmm} have
typically written scripts in a text editor and run the scripts by passing them to the simulator on the command line.
The simulator in turn writes the results to disk and the user analyzes the results
by reading the files manually or passing them to other programs.

With the arrival of software like Rasmol\cite{sayle1995rasmol} and VMD\cite{Humphrey1996Vmd},
it became easy to both visualize the trajectories and analyze physical quantities such as the radial distribution function.
VMD came with support both for reading trajectories from file and attaching to a
running process using MDComm\cite{nelson1995mdscope} and a supported simulator such as NAMD\cite{nelson1996namd}.
Although this enabled real-time visualization, the workflow still involved using all these different programs which is still a problem today.

Analysis of simulations is often done as a postprocessing step where the trajectories and other per-atom quantities (velocity, stress etc.)
are loaded into an analysis tool written by researchers themselves. About 10 years ago, a new visualization tool called OVITO was released \cite{Stukowski2009Visualization},
with a powerful modifier pipeline where input data can be processed and manipulated as a series of building blocks.
These building blocks can be used to calculate new properties and atoms can be colored thereafter.

Scalar quantities such as temperature, pressure and number of atoms in a specific state are often printed to a
text file. Researchers often write their own plotting scripts in languages like MATLAB or Python
to see how these quantities evolve with time.

@@

It is especially useful during development of new scripts, but can be used
to run full, publication-ready simulations with minimal simulation overhead.
The current implementation only supports LAMMPS as a simulator, but is written in a general
way so integrating other codes is a simple task.

@@

\section{\label{sec:features}Summary of features}
Atomify was designed to improve the workflow as its main goal.
Focusing on minimizing the number of context switches and mouse clicks reduces
the overhead between creativity and immediate results from a simulation.
It has a script editor, visualization window and a summary panel so all relevant parts
of a simulation can be on the screen at the same time.

Any changes to the script like defining regions or groups,
testing temperatures, size of simulation box or inserting atoms can
quickly be verified since the visualization shows the real-time state of the simulator.
Rerunning the current script can be done by pressing \keys{Ctrl+R} (\keys{cmd+R} on macOS).

How the pressure, radial distribution function or some other
complicated variable evolves over time is easily understood
since these can be plotted real-time with a simple click in the list of variables.
Bonds between atoms in a molecule is automatically extracted from LAMMPS, or can
be defined as a distance threshold between atom types.
Regions and groups can be highlighted or hidden which is useful to focus on some parts of a system.
Rendering periodic copies can be done to study features near a periodic boundary condition.

Per-atom scalar quantities can be used to color each atom based its value.
For instance, \code{compute cna/atom}\cite{faken1994systematic, tsuzuki2007structural} can be used to identify the crystal structure an atom is a part of.
It gives each atom an integer value in the range [1,5] representing FCC, HCP, BCC, Icosohedral or unknown crystal structure.
\code{compute displace/atom} can be used to see how far each atom has moved and \code{compute stress/atom} gives the local stress in the system.
These values can be used as input to \code{fix ave/atom} to get time averaged values to reduce noise.
Coloring each atom based on these values can give great insight when it is available with a single mouse click.

Although LAMMPS is mostly used to study molecular dynamics, it can also be used to
study monte carlo simulations\cite{frenkel2001understanding} and granular
materials\cite{brilliantov1996model, silbert2001granular, zhang2005jamming},
both of which work very well in Atomify since they are particle based simulations.

Atomify supports rendering of spheres and cylinders representing atoms and bonds.
It is done using the powerful Qt3D engine with custom visualization techniques such as
billboard raytracing\cite{tarini2006ambient} and screen space ambient occlusion\cite{bavoil2008screen}
for improved dynamic depth perception. This allows a large amount of atoms and molecules being rendered with excellent framerate.
The liming factor is the simulation itself since force calculation is a heavy task.

The Atomify GUI is shown in figure \ref{fig:gui} where \todo{short summary of GUI}.
This turns out to improve the workflow substantially.
In addition, Atomify comes with multiple examples,
many of which have been made by members of the LAMMPS community.

\section{\label{sec:casestudy}Case study: flow through a narrow nanotube}
\todo[inline]{More references in here?}
The features of Atomify is best demonstrated through a case study where many of
them are useful to obtain a productive workflow during script development.
We have chosen to create a system with fluid flow inside a tight nanochannel.
This involves creating a somewhat complicated initial geometry with multiple regions
where we apply different rules such as a dissipative thermostat, constant force mimicking
applied pressure gradient, and frozen atoms to avoid net momentum of the solid.
We will create two different atom types so we can have a gas phase and a solid phase at the same temperature.

@@

\section{Availability}
Atomify is available on Linux and MacOS and can be downloaded from
\href{https://ovilab.net/atomify}{http://ovilab.net/atomify}.
A lightweight version of Atomify is also available for mobile devices running
Android or iOS.

\todo[inline]{Potentially future work, emscripten, webassembly? emscripten has a proceedings paper.}
